---
title: "Gentrification after a Natural Disaster: The Case of Hurricane Katrina"
output: 
  html_document:
    css: bodycss.css
    fig_width: 8
    font-family: Helvetica,Arial,sans-serif;
    number_section: yes
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

<style type="text/css">
body{ 
      font-size: 15px;
      font-family: Helvetica,Arial,sans-serif;
      line-height: 200%;
  }
  
.author {
 font-size: 15px;
 color: Black;
 font-style: normal;
 text-align: center;
}


.date { 
 font-size: 15px;
 color: Black;
 font-style: normal;
 text-align: center;
}

.title{
  text-align: center;
  font-size: 15px;
 color: Black;
 
}

.toc-ignore{
  text-align: center;
  font-size: 15px;
 color: Black;
}

.fluid-row{
  text-align: center;
  font-size: 15px;
 color: Black;
}

</style>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE,echo=FALSE)
```


```{r init_mapdata}
rm(list=ls())
library(rgdal)
library(ggplot2)
library(rgeos)
library(proj4)
library(ggmap)
library(sp)
library(plyr)
library(data.table)
library(lfe)
library(stargazer)


pointsinside <- function(point,shapefile) {
  output <- rep(NA,nrow(point))
  pb <- txtProgressBar(min = 1, max = nrow(point), style = 3)
  for(i in 1:nrow(point)) {
      setTxtProgressBar(pb, i)
      temp   <- SpatialPoints(point[i,],proj4string=CRS(proj4string(shapefile)))
      output[i] <- gContains(shapefile,temp)
  }
  return (output)
}

ms_flood_raw <- readOGR("E:/gentrification/Katrina Flood Maps","Katrina_MS_SurgeInundationLimit")#  katrina_ms_abfe_contours
ms_flood_raw <- gUnaryUnion(ms_flood_raw)
proj4string <- "+proj=tmerc +lat_0=29.5 +lon_0=-88.83333333333333 +k=0.99995 +x_0=300000 +y_0=0 +datum=NAD83 +units=us-ft +no_defs +ellps=GRS80 +towgs84=0,0,0"
ms_flood <- fortify(ms_flood_raw)
xy <- data.frame(x=ms_flood$long, y=ms_flood$lat)
pj <- project(xy, proj4string, inverse=TRUE)
ms_flood$long <- pj$x
ms_flood$lat <- pj$y

# cencustracts <- readOGR("E:/gentrification/CensusTracts Maps","cb_2017_28_tract_500k") 
# cencustracts <- cencustracts[cencustracts$COUNTYFP %in% c("047","045","059"),]
# cencustracts.df <- as(cencustracts,"data.frame")
# cencustracts <- fortify(cencustracts,region="GEOID")
# cencustracts <- merge(cencustracts,cencustracts.df,by.x='id',by.y='GEOID')
#   dat <- cencustracts
#   pj <- project(data.frame(x=dat$long,y=dat$lat), proj4string, inverse=FALSE)
#   dat$long <- pj$x
#   dat$lat <- pj$y
#   dat <- dat[,c("long","lat")]
# cencustracts['inside_flood'] <- as.numeric(pointsinside(dat,ms_flood_raw))
# save(cencustracts,file="E:/HarveyData/CensusTracts/cencustracts_ms.rda")
load(file="E:/gentrification/CensusTracts Maps/cencustracts_ms.rda")


insidecount <- ddply(cencustracts,.(id),summarise,inside=sum(inside_flood),count=length(inside_flood))
insidecount['inside_pct'] <- insidecount$inside/insidecount$count
insidecount <- insidecount[,c("id","inside_pct")]
cencustracts <- merge(cencustracts,insidecount,by="id")
cencustracts$TRACTCE <- as.numeric(substr(floor(as.numeric(cencustracts$id)/100)*100,6,25))
cencustracts$COUNTYFP <- as.numeric(as.character(cencustracts$COUNTYFP))
cencustracts['treat'] <- ifelse(cencustracts$inside_pct>=0.8,1,ifelse(cencustracts$inside_pct<=0.2,-1,0))
cencustracts['county_tract'] <- paste(cencustracts$COUNTYFP,cencustracts$TRACTCE)

cencustracts_treat <- unique(cencustracts[cencustracts$inside_pct>=0.8,]$county_tract)
cencustracts_treat <- c(cencustracts_treat,"47 600", "47 1300","47 100","47 2800","59 40600","59 41000","47 3600")

cencustracts_ctrl <- unique(cencustracts[cencustracts$inside_pct<=0.2,]$county_tract) # [9800]    
cencustracts_ctrl <- c(cencustracts_ctrl,"47 1700","47 2400","47 2700","59 40800","59 40700","59 40400","47 3400" )

```


```{r init_hmda_data}
files = list.files(path = 'E:/gentrification/HMDA MS Katrina/2000-2003', pattern = '^MS_045047059')
dat_list = lapply(paste('E:/gentrification/HMDA MS Katrina/2000-2003/',files,sep = ""), function (x) data.table(readRDS(x)))
hmda_2000_2003 = rbindlist(dat_list, fill = TRUE)
hmda_2000_2003$censustract <- floor((as.numeric(hmda_2000_2003$censustract)*100)/100)*100
hmda_2000_2003 <- as.data.frame(hmda_2000_2003)
hmda_2000_2003['county_tract'] <- paste(hmda_2000_2003$countycode,hmda_2000_2003$censustract)


files = list.files(path = 'E:/gentrification/HMDA MS Katrina/2004-2010', pattern = '^MS_045047059')
dat_list = lapply(paste('E:/gentrification/HMDA MS Katrina/2004-2010/',files,sep = ""), function (x) data.table(readRDS(x)))
hmda_2004_2010 = rbindlist(dat_list, fill = TRUE)
hmda_2004_2010$censustract <- floor((as.numeric(hmda_2004_2010$censustract)*100)/100)*100
hmda_2004_2010 <- as.data.frame(hmda_2004_2010)
hmda_2004_2010['county_tract'] <- paste(hmda_2004_2010$countycode,hmda_2004_2010$censustract)

hmda_2004_2010 <- hmda_2004_2010[,names(hmda_2004_2010) %in% names(hmda_2000_2003)]

hmda <- rbind(hmda_2000_2003,hmda_2004_2010)

hmda <- hmda[hmda$county_tract %in% c(cencustracts_ctrl,cencustracts_treat),]
hmda <- hmda[!hmda$censustract %in% c("47 1500","47 3200"),] # in both treatment and control

hmda['post'] <- ifelse(hmda$asofdate>=2006,1,0)
hmda['treat'] <- ifelse(hmda$county_tract %in% cencustracts_treat, 1,0)

hmda['nonwhite'] <- ifelse(hmda$applicantrace1 != 5,1,0)
hmda['primaryfemale'] <- ifelse(hmda$applicantsex==2,1,0)
hmda['approved'] <- ifelse(hmda$actiontaken<=2,1,0)
hmda['jointapplication'] <- ifelse(hmda$coapplicantsex!=5,1,0)
hmda$applicantincome <- as.numeric(hmda$applicantincome)

hmda['nonwhitenonasian'] <- ifelse(!hmda$applicantrace1 %in% c(5,2),1,0)

# hmda$ratespread <- as.numeric(hmda$ratespread)
```

```{r flood_map}
googlemap = get_map(location = c(-89, 30.4), zoom = 9, source = "google", maptype="terrain")

cnames <- aggregate(cbind(long, lat) ~ TRACTCE, data=cencustracts, FUN=median)

# map <- ggmap(googlemap)+
map <- ggplot(data=ms_flood,aes(x=long,y=lat,group=group))+
  geom_polygon(fill='red',alpha=0.5)+
  scale_x_continuous( limits = c( -89.7 , -88.3 ) )+
  scale_y_continuous( limits = c( 30.1 , 30.7 ) )+
  geom_path(aes(x=long,y=lat,group=group),data=cencustracts,color = 'red',size=0.3)+
  geom_path(aes(x=long,y=lat,group=group),data=cencustracts[cencustracts$treat==1,],color = 'blue',size=1)+
  geom_path(aes(x=long,y=lat,group=group),data=cencustracts[cencustracts$treat==-1,],color = 'gray40',size=1)#+
  #geom_text(data=cnames, aes( fontface=1.5 ,cnames$long, cnames$lat , label = cnames$TRACTCE,group=TRACTCE), color= "black"  )

map

# boundary<-  ggplot(data=cencustracts,aes(x=long,y=lat,group=group))+geom_path()+geom_text(data=cnames, aes( fontface=1 ,cnames$long, cnames$lat , label = cnames$TRACTCE,group=TRACTCE), color= "black",size=2 )  
# boundary
```


```{r formulas}
applicantincome = "log(applicantincome)~post*treat+nonwhite+primaryfemale+jointapplication|asofdate+countycode|0|asofdate+county_tract"
nonwhite = "nonwhite~post*treat+log(applicantincome)+primaryfemale+jointapplication|asofdate+countycode|0|asofdate+county_tract"
primaryfemale ="primaryfemale~post*treat+log(applicantincome)+nonwhite+jointapplication|asofdate+countycode|0|asofdate+county_tract"
approved = "approved~post*treat+log(applicantincome)+nonwhite+jointapplication+primaryfemale|asofdate+countycode|0|asofdate+county_tract"

dep.var.labels = c("log(applicantincome)","nonwhite","primaryfemale","approved")
```


# New Purchases, Owner occupied
```{r}

sample <- hmda[hmda$purposeofloan==1 & hmda$occupancy==1 ,] # new purchases, owner occupied

regs <- list()
regs[[1]] <- felm(as.formula(applicantincome),data = sample)
regs[[2]] <- felm(as.formula(nonwhite),data = sample[sample$applicantrace1!=6,])
regs[[3]] <- felm(as.formula(primaryfemale),data = sample[sample$jointapplication==0,])
regs[[4]] <- felm(as.formula(approved),data = sample)

stargazer(regs,type="text",omit.stat=c("ser"),dep.var.labels = dep.var.labels)
```


# New Purchases, Non-owner occupied
```{r}

sample <- hmda[hmda$purposeofloan!=3 & hmda$occupancy!=1 ,] 

regs <- list()
regs[[1]] <- felm(as.formula(applicantincome),data = sample)
regs[[2]] <- felm(as.formula(nonwhite),data = sample[sample$applicantrace1!=6,])
regs[[3]] <- felm(as.formula(primaryfemale),data = sample[sample$jointapplication==0,])
regs[[4]] <- felm(as.formula(approved),data = sample)

stargazer(regs,type="text",omit.stat=c("ser"),dep.var.labels = dep.var.labels)
```


# Refinances
```{r}

sample <- hmda[hmda$purposeofloan==3,] 

regs <- list()
regs[[1]] <- felm(as.formula(applicantincome),data = sample)
regs[[2]] <- felm(as.formula(nonwhite),data = sample[sample$applicantrace1!=6,])
regs[[3]] <- felm(as.formula(primaryfemale),data = sample[sample$jointapplication==0,])
regs[[4]] <- felm(as.formula(approved),data = sample)


stargazer(regs,type="text",omit.stat=c("ser"),dep.var.labels = dep.var.labels)
```

